Deep Reinforcement Learning for Conservation Decisions

### 2021-05-05 manuscript/manuscript.Rmd
some minor comments on intro

<!-- Environmental problems, ML -->
Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
But will ML approaches really help us tackle a changing planet?
Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* (e.g. image recognition) and *unsupervised learning* (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:

@@

Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
RL agents have recently achieved superhuman performance on classic Atari games [@DQN], while Deepmind’s RL-trained AlphaZero has defeated the world's grandmasters in Go, Chess, and Shogi [@alphazero].
Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
Of course real ecosystems are also far more complex than a game of chess or MoonLander -- but can the same be said of our efforts at finding the best strategy?
Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management. 
Even simpler models are frequently employed throughout ecology to guide understanding and response.
Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
We believe that the combination of such approaches from RL trained using sufficient simulations of suitable ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.

In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.
Our second example considers ecosystem management under slow environmental degradation: a scenario which is not amenable to classic optimization methods such as dynamic programming.
In this case there is no existing optimal solution to compare against, but an appropriately tuned RL agent can out-perform a sensible rule-of-thumb response.
In our third example simulating wildfire spread highlights an open problem which we expect will be considerably harder to find effective solutions.
These examples demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
We further include implementations of three fully featured simulation modules following current standards for RL benchmark problems as Python modules distributed on PyPi and GitHub.
Each module includes several more challenging variations to the examples described here.
We intend to continue to expand this library of ecological management benchmarks with the hope both of challenging researchers already working in RL to tackle some of these pressing issues and for ecologists and environmental scientists to continue to refine these benchmarks into more realistic ecological simulations.

The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.
Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.

@@

>>>>

<!-- Environmental problems, ML -->
Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].

<!-- citation suggestion for declining biodiversity  -->

<!-- @article{dirzo2014defaunation, -->
<!--   title={Defaunation in the Anthropocene}, -->
<!--   author={Dirzo, Rodolfo and Young, Hillary S and Galetti, Mauro and Ceballos, Gerardo and Isaac, Nick JB and Collen, Ben}, -->
<!--   journal={science}, -->
<!--   volume={345}, -->
<!--   number={6195}, -->
<!--   pages={401--406}, -->
<!--   year={2014}, -->
<!--   publisher={American Association for the Advancement of Science} -->
<!-- } -->

But will ML approaches really help us tackle a changing planet?

<!-- suggestion: "...tackle the most pressing environmental challenges on a changing planet" -->

Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* (e.g. image recognition) and *unsupervised learning* (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:

@@

<!-- really like this list! super clear and helpful! -->

Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
RL agents have recently achieved superhuman performance on classic Atari games [@DQN], while Deepmind’s RL-trained AlphaZero has defeated the world's grandmasters in Go, Chess, and Shogi [@alphazero].
Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
Of course real ecosystems are also far more complex than a game of chess or MoonLander -- but can the same be said of our efforts at finding the best strategy?
Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.

<!-- another widely used model is ED2 (ecosystem demography model) - I think more common than SORITE? https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2008JG000812 -->

Even simpler models are frequently employed throughout ecology to guide understanding and response.
Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].

<!-- I think this is missing the connection between the line 88 examples and needing more sophisticated methods for decision making below. i.e. "As we develop more complex process based models (such as Ecosim and ED2), methods with the capacity to effectively integrate high dimensional models into decision making processes will be critical to ensuring their policy relevance" (or something like that, then go on to NN ML)-->

Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
We believe that the combination of such approaches from RL trained using sufficient simulations of suitable ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.

In this paper, we draw on examples from fisheries management, ecological tipping points, and wildfire spread to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.

<!-- ^ awesome! -->

Our simplest example approaches the known optimal solution for a classic fisheries management problem with no prior knowledge of the underlying model.

Our second example considers ecosystem management under slow environmental degradation: a scenario which is not amenable to classic optimization methods such as dynamic programming.
In this case there is no existing optimal solution to compare against, but an appropriately tuned RL agent can out-perform a sensible rule-of-thumb response.
In our third example simulating wildfire spread highlights an open problem which we expect will be considerably harder to find effective solutions.

 <!-- I think this is pretty specific (almost like a discussion paragraph, but we haven't read the paper yet.) Could you be more broad about what you set out to do in the paper? i.e.: "We start with a simple example from fisheries management that has a known optimal solution, proving that with no prior knowledge of underlying model dynamics RL agents can find a close to optimal solution. We go on to show that RL can outperform rule-of-thumb solutions to management scenarios not amenable to classic optimization methods such as dynamic programming. Our examples demonstrate that..."  -->

These examples demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.
We further include implementations of three fully featured simulation modules following current standards for RL benchmark problems as Python modules distributed on PyPi and GitHub.
Each module includes several more challenging variations to the examples described here.
We intend to continue to expand this library of ecological management benchmarks with the hope both of challenging researchers already working in RL to tackle some of these pressing issues and for ecologists and environmental scientists to continue to refine these benchmarks into more realistic ecological simulations.

The pitfalls of RL-based solutions to global environmental change problems are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary.
We discuss not only technical issues that can cause RL solutions to fail in surprising ways, but also potential issues raised by the adoption of more algorithmic approaches to conservation generally.
We emphasize how an open, transparent and reproducible approach can help mitigate some (but not all) of these concerns, and facilitate a more effective interface between teams of researchers from both ecological and computer sciences.
If algorithms are to play a meaningful role in the allocation and exploitation of natural resources from fish to forests, good intentions [@Joppa] may not be sufficient to ensure equitable outcomes.

<!-- maybe cite Caleb's paper here? -->

Ecologists and computer scientists will also need input from experts in governance and ethics to ensure that applications of AI to ecological challenges are fair and just.

<!-- these last few sentences almost feel like discussion points. I'm torn though, because it it nice to have them in the intro-->

@@

<!-- something to introduce this section would be helpful for transition: "We start with a brief overview of RL problems and solutions:" -->
